---
title: "Challenges"
author: Nick Ulle
date: August 11, 2016
output: github_document
---

## Inferring Call Types

Suppose a function's return type depends on its input type:

```{r, eval=FALSE}
bar = function(x) {
  if (class(x) == "complex")
    return (42+0i)

  return (42L)
}

x = foo()
y = bar(x)
stopifnot(is(y, "integer"))
```

In this example, the type of `y` cannot be determined until after the type of
`x` is determined. The subsequent assertion that `y` will have type `integer`
doesn't guarantee the type of `y` before that line: it's possible the
programmer made a mistake when assigning `x`, and type inference must be able
to detect whether or not the assertion will pass. Indeed, inference should
alert the programmer before run-time if the assertion will fail.

A naive approach to this problem is to solve the partial constraint set
immediately when the calls `foo` and `bar` are encountered. However, if
constraints from subsequent expressions are necessary to infer the type of `x`,
this is unlikely to work well [TODO: Examples?].

A second approach is to lift the predicates that determine the types of the
calls from the function bodies; they can then be inserted in an `IfType`. The
resulting constraints would look like this:

```
x#1 ~ RealType
y#1 ~ IfType(class(x) == "complex", ComplexType, IntegerType)
y#1 ~ IntegerType
```

This has the benefit of making it easy to identify the run-time predicates that
affect the constraints for the entire program. However, it strips the
predicates of context; the `x` referred to in the predicate corresponds to a
type variable for `foo`'s type constraints rather than the top-level
constraints.

Another option is to assign a type variable to the calls as if they were
symbols. This leads to the system:

```
x#1 ~ TypeOf foo()
y#1 ~ TypeOf bar(x#1)
y#1 ~ IntegerType
```

This keeps type detection and inference separate, since the system can be
solved _after_ detection. The system can be solved with an iterative procedure
that unifies constraints and then infers call types.


## If Statements

### Value Predicates

```{r, eval=FALSE}
x = if (sample(0:1, 1)) "Text" else 8L
```

This could generate the constraint:

```
x#1 ~ IfType(sample(0:1, 1), StringType, IntegerType)
```

In general, these prevent inference from succeeding.

### Type Predicates

A special case of the example above occurs when the predicate depends only on
type information:

```{r, eval=FALSE}
x = if (is(y, "character")) "Text" else 8L
```

### Other Predicates

Are there any other cases where it's likely the predicate can be resolved
before run-time? Technically, such predicates should be optimized out of the
program anyways: they are useless tests.

## Loops

Suppose the iterator variable in a loop takes a different type on every
iteration:

```{r, eval=FALSE}
my_list = list(TRUE, 1L, 1.0, 1.0+0i, "1.0")

for (x in my_list) {
  # ...
}
```

This example poses several challenges. First, type inference must identify the
type of `x` for every iteration, which means a type constraint must be
generated for every iteration. Moreover, the data structure for the results of
inference must accommodate this.

Second, unless the loop body casts `x` to a common type (which may be
implicit), the compiler must emit a block for every iteration.
