---
title: "Type Constraints"
author: Nick Ulle
date: August 16, 2016
output: github_document
---

This document examines how type constraints should be generated for
assignments.

Every assignment must have a symbol as its write operand [is this true for call
assignments?]. Here we'll consider different cases of read operands. In all
cases, it's assumed that assignment changes the type of the write symbol, so a
new type variable is bound to that symbol and marked as current reaching type.

## Literals

```{r}
x = 3.0
```
Literals are the most straightforward case because their type is known. A new
constraint is added to the system with the literal's type.

Assuming `x` previously had type `x1`:

1. The write operand is bound to `x2`.
2. The reaching type of `x` is set to `x2`.
3. The constraint `x2 <=> RealType` is added to the constraint set.

## Symbols

```{r}
x = y
```
Symbols are not much more complicated than literals. The write operand is
handled as described for literals. The read operand is not mutated, so its type
variable does not need to be changed. The constraint $x2 ~ y1$ is added to the
constraint set.

[What should happen if `y` has no reaching type?]

## Binary Operations & Calls

```{r}
x = 3.0 + 4.1
```
The return type of the call is resolved immediately by the upcasting rules.

### Unresolved Types

```{r}
x = 3.0 + y
```

Assuming `x`, `y` previously had types `x1`, `y1`:

1. The write operand is bound to `x2`.
2. The reaching type of `x` is set to `x2`.
3. The constraint `x2 <=> UnresolvedCall("+", RealType, y1)` is added to the
   constraint set.

This defers inference for the return type of `+` until it's demanded by the
unification algorithm. When the type is needed, (the hope is that) the type of
`y1` will already have been inferred.

We may also need to capture the environment the call was made in. For instance,
consider this example:

```{r}
sum = function(...) "Hi, I'm sum!"

sum(4, 3)
```

Overrides for functions may also occur in an ancestor of the inference
environment, and these must be found and handled correctly. Thus lookup for
calls needs to proceed from the environment associated with the code.
