---
title: "If-Statement Examples"
author: Nick Ulle
date: August 18, 2016
output: github_document
---

```{r}
x = if (y) "Text" else 8L

# or equivalently:

if (a)
  x = 4+2i
else
  x = 8L
```

This example is the simplest case of types changing due to an if-statement.
Here we will focus on the second form, because the first form can always be
rewritten as the second, and the second does not emphasize any one assignment
(there may be several assignments in the body of an if-statement).

Assuming that `x` previously had type `x#1`:

1. The write operand in the `if` branch is bound to `x#2`.
2. The reaching type of `x` is set to `x#2`.
3. The constraint `x#2 <=> ComplexType` is added.

4. The write operand in the `else` branch is bound to `x#3`.
5. The reaching type of `x` is set to `x#3`.
6. The constraint `x#3 <=> IntegerType` is added.

7. At the end of the if-statement, the reaching type of `x` is set to `x#4`.
8. The constraint `x#4 <=> RuntimeType(x#2, x#3, node)` is added, where `node`
   is a reference to the `If` node in the AST.

In most cases, there will be more than one assignment. For example:
```{r}
if (a) {
  x = 4+2i
  y = 12.0
} else {
  x = 8L
  y = 12L
}
```

The steps are similar to above, but must be carried out for both `x` and `y`.
