---
title: "Type Resolution"
author: Nick Ulle
date: August 30, 2016
output: github_document
---

## Literal Propagation

Consider the simple system
```
x#1 <=> IntegerType
y#1 <=> x#1
```
Solving this system proceeds as follows:

1. Unify `x#1` and `IntegerType`; the substitution `x#1 => IntegerType` is
   added to the solution set.
2. Apply the solution set to the remaining constraints, which yields the system
   ```
   Soln:
   x#1 => IntegerType
   Cons:
   y#1 <=> IntegerType
   ```
3. Repeat steps 1-2.

What if a constraint is encountered that can't be solved? For example:
```
y#1 <=> x#1
x#1 <=> IntegerType
```
This case shows the need to apply the new solution to the existing solution set
at each step. That is,

1. Unify `x#1` and `y#1`. Apply the new solution to both the constraint set and
   the solution set, so:
   ```
   Soln:
   y#1 => x#1
   Cons:
   x#1 <=> IntegerType
   ```
2. Unify `x#1` and `IntegerType`. Apply the new solution to both the contraint
   set and the solution set, so:
   ```
   Soln:
   y#1 => IntegerType
   x#1 => IntegerType
   ```

## Unresolved Calls

Consider this simple example:
```
y#1 <=> UnresolvedCall(length, x = x#1)
```
The `length` function always returns an integer, so the type of `x#1` is not
even needed. Resolution should produce this solution:
```
y#1 => IntegerType
```

```{r}
x = 3i
y = 4 + x
```
The constraint set is:
```
x#1 <=> ComplexType
y#1 <=> UnresolvedCall(+, RealType, x#1)
```
The solution set is:
```
x#1 => ComplexType
y#1 => ComplexType
```

What should happen if one of the constraints has an as yet unknown type? And
how can this happen in examples?
```{r}
# Assume x is a parameter.
y = 4 + x
```

