---
title: "Fibonacci Vignette"
author: "Nick Ulle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fibonacci Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Fibonacci Vignette

```{r}
library(RTypeInference)
```

This vignette demonstrates running type inference on a simple, non-recursive 
function, `fib()`, for calculating the n-th Fibonacci number:

```{r}
fib = function(n) {
  .typeInfo(n = IntegerType())

  acc = c(0, 0, 1)
  for (i in 1:n) {
    acc[1:2] = acc[2:3]
    acc[3] = sum(acc[1:2])
  }

  return(acc[1])
}
```

The function `.typeInfo()` is provided by `RTypeInference` for annotating 
argument types; it's defined as a no-op so that functions using it can still be 
run in the R interpreter with minimal performance overhead. In the future, we 
will also support annotating argument types using an attribute.

```{r}
sapply(1:10, fib)
```

Two important details are that no type information is provided about the 
variable `acc`, and the function returns a subset of this variable. Moreover, 
the type of the variable `i` in the loop should not only be inferred, but also 
marked as an iterator variable.

In order to hold on to the inferred types, we must first define a 
`TypeCollector`:

```{r}
collector = TypeCollector()
```

Now we can infer the types of the function:

```{r}
inferTypes(fib, collector)
```

Although the vector `acc` has numeric type in the R interpreter, its type is 
inferred as integer because the initial values are all integers. Thus the 
return type is inferred correctly. Note that the return value is marked as 
unknown; for a constant function, the return value would be displayed.

To view the inferred types for the other variables, we have to print the 
`TypeCollector` object we defined earlier:

```{r}
print(collector)
```

In the printout we can see that the inferred types for both `acc` and `i` are 
correct. In particular, the variable `i` is marked as an iterator variable with 
atomic type integer.

