---
title: "RTypeInference Model"
author: "Nick Ulle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide"}
library(RTypeInference)
```

## `infer_types()`

To get the types for the variables in an R program, we use the `infer_types()`
function. The input is the control flow graph for the R program, in SSA form.
Computing the CFG is handled by the __rstatic__ package, and the details are
discussed in that package's documentation. Here we will take those details for
granted.

The idea of type inference is to generate a system of constraints on the types
used in the program, and then solve the system to get the types. The
`infer_types()` function performs both of these steps, but they can also be
carried out separately with `constrain()` and `solve()`, respectively.

Each constraint comes from a different piece of evidence in the code about the
types. For instance, an assignment `x = "hi"` signals that the variable `x` is
a string. Many expressions allow us to narrow down the set of valid types for a
variable, but don't specify the exact type. When we see `x + y`, we know that
unless `+` has been redefined by the user, `x` and `y` must both be numbers.
However, we cannot tell from this expression alone whether they are logical
values, integers, floating point numbers, or complex numbers.

R allows variables to be redefined with a different type at any point in a
program, so most variables are associated with a sequence of types rather than
a single type. Knowing the sequence of types for a variable is only useful if
we also know where in the program each type occurs.

To deal with this problem, constraints always refer to single static assignment
(SSA) names rather than variable names. In SSA form, each (re)definition of a
variable is given a unique name. Because of this, each SSA name can be
unambiguously associated with a single type in a variable's sequence of types.

SSA names are not the only strategy for keeping track of constraints and types.
We could think of types as being associated with specific expressions within a
program rather than variables. This is a conceptually different approach to the
problem. In this approach, type inference must find a type for every expression
in the program. This is not any more difficult, but does require additional
bookkeeping with associated memory costs.

In order to generate constraints, `constrain()` traverses the program's CFG and
calls the S3 generic `constrain_ast()` on each expression. For now, we can
customize how constraints are generated for arbitrary expressions by redefining
the `constrain_ast()` methods as needed.

<!-- FIXME: Need a better example! -->
Consider a function that finds the area of an annulus when given the inner and
outer radius

```{r}
circle_area = function(r_inner, r_outer) {
  outer = pi * r_outer^2
  inner = pi * r_inner^2

  return (outer - inner)
}
```

We can compute the constraints for this with

```{r}
cfg = rstatic::to_blocks(circle_area)

cons = constrain(cfg)
```

Constraints are collected in a `ConstraintSet`, which has reference semantics
to simplify writing recursive `constrain_ast()` methods. The order of the
constraints within the set does not matter. This is important because it allows
us to set constraints based on how a variable is used after the point where
that variable was defined.

In the example, we can see that the right-hand side of each constraint is
printed like a function call. For now, types are not computed for calls during
constraint generation; the computation is deferred until the system of
constraints is solved.

By using `constrain()` and `solve()` separately, we can add to and modify the
constraints before solving. Constraints can be accessed by index with `[[`
(although they cannot yet be assigned with `[[`). The method 

---

The returned `SolutionSet` lists the inferred type of each SSA name defined in
the function.


<!-- FIXME: OTHER STUFF
The `add` function is _polymorphic_: the arguments are not constrained to any
particular type. In other words, `add(3, 4)`, where both arguments are
integers, will work just as well as `add(3.1, 5+1i)`, where the arguments are a
real and a complex number.
-->


