---
title: "RTypeInference"
author: "Nick Ulle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Using RTypeInference

The **RTypeInference** package provides type inference for R code.

The type inference strategy is based on Damas-Milner type inference. This
strategy collects information about the type of each expression in the source
code, combining information about related expressions.

As an example, consider the expression `x <- 3.14`. The value `3.14` has type
double in R.  Since the expression assigns `3.14` to the variable `x`, we can
conclude that the variable `x` must also have type `numeric`. 

Let's try out **RTypeInference** for this example.

First, we need to use **rstatic**'s `quote_blocks` function to quote the
expression. The `quote_blocks` function converts the expression into static
single-assignment (SSA) form. A key feature of SSA form is that it numbers each
variable assignment, so that different definitions of the same variable have
distinct names. This is convenient for type inference because R allows
variables to be assigned different types at different points in the source
code.

Second, we call **RTypeInference**'s `constrain` function on the quoted
expression in order to collect information about the types.

Here's the complete code for the two steps described:
```{r, message = FALSE}
library(rstatic)
library(RTypeInference)
```

```{r}
ex = quote_blocks(x <- 3.14)
constrain(ex)
```

The `constrain` function returns a `Result` object, an S4 object that
represents intermediate results in **RTypeInference**. The object has two
slots. We'll examine the first slot, `constraints` in a later example. The
second slot, `map`, contains a mapping between variables in the source code and
types.

The mapping reports that the variable `x_1` is assigned ("defined as") a
numeric value. Here `x_1` corresponds to the variable `x` in the original
expression. The `_1` is the number added by the SSA form.

The mapping also reports that the variable `x_1` is used (that is, read from)
later in the code. This is a limitation of the current version of
**RTypeInference**, likely to be fixed in later versions. For the curious
reader, here's what's happening: **RTypeInference** assumes that any source
code passed in for type inference is part of a function definition. The SSA
form makes R's implicit return rule for functions explicit by inserting a
variable `._return_` into the code and assigning the last expression to
`._return_`. Thus type inference processes the assignment `x <- 3.14` and also
the assignment `._return_ <- x`. That's why there's an extra entry for
`._return_` in the mapping.

The mapping is an S4 object of class `SymbolMap`. The class inherits `list`,
and uses one named element for each variable in the source code.

Let's extend the example with another variable. Consider this code:
```{r, eval = FALSE}
x <- 3.14
y <- x
```

The steps to run inference on this code are the same as before. Since the code
has multiple lines, it needs to be surrounded by curly braces when passed to
`quote_blocks`:

```{r}
ex = quote_blocks({
  x <- 3.14
  y <- x
})
ans = constrain(ex)
ans
```

Let's look at how the results correspond to the code. As before, the map
reports that `x_1` is defined as `RNumeric`. The map also reports that the type
of `x_1` is used somewhere, as `t1`. The `t1` is a placeholder, called a _type
variable_, for the type of `x_1`. Type variables provide the type inference
system with three kinds of flexibility:

1. _As placeholders for types that haven't been inferred yet._ For instance, if
   the type of `x` was not clear from the assignment in the code above, we
   might still want to record the fact that `y` has the same type as `x`. These
   type variables are the ones type inference ultimately tries to "solve" for,
   replacing them with concrete types.

2. _As indicators of polymorphism._ A polymorphic function is one that accepts
   more than one type of argument. In Damas-Milner type systems, polymorphism
   is all or nothing: a polymorphic function must be able to accept any type,
   and type variables are used to indicate polymorphism. Type variables that
   indicate polymorphism remain even after type inference is complete. In R,
   polymorphic functions are often restricted to a specific set of types rather
   than accepting any type.

3. _As indirection that makes the type inference system more robust._
   **RTypeInference** generates a new type variable each time a program
   variable is _used_, even if an earlier assignment establishes the type of
   that program variable. By not propagating the type information gathered from
   the assignment immediately, the step in type inference that collects
   information from source code is decoupled from the step in type inference
   that "solves" for the types by propagating the collected information. As a
   result, the information collection step always analyzes the entire source
   code, even if the source code contains type errors.

A _constraint_ is a relationship between a pair of types (including type
variables). Through the `constrain` function, type inference collects
constraints based on the expressions in the source code. These appear in the
`constraints` slot of the result. 

In the example code, the assignment `y_1 = x_1` causes type inference to
generate a type variable `t1` as a placeholder for the type of `x_1`. Since the
earlier assignment `x_1 = 3.14` reveals that the type of `x_1` is numeric, the
type inference system also generates a constraint that `t1` is equivalent to
the type `RNumeric`. This constraint is the first element of the `constraints`
slot. Let's examine the constraint: 
```{r}
ans@constraints[[1]]
```

Each constraint records a relationship between a pair of types; these two
elements correspond to slots `t1` and `t2` (no relation the type variable `t1`)
on the constraint object. The **typesys** class `Equivalence` represents an
equivalence constraint: the two types must be the same. Here we can see that
`t1` and `t2` contain the type variable `t1` and the type `RNumeric`,
respectively.

Constraints also have a slot `src` with a reference to the source code
expression that caused the constraint to be generated. In this case, the source
code expression that caused the constraint to be generated is the expression
`x_1`, which is part of the assignment `y_1 <- x_1`:
```{r}
con1 = ans@constraints[[1]]
con1@src
con1@src$parent
```

As mentioned earlier, **RTypeInference** assumes the source code being analyzed
is part of a function definition, so we also see constraints related to the
return value (constraints 2 and 3 here). For instance, consider the second
constraint:
```{r}
con2 = ans@constraints[[2]]
con2

con2@src$parent
```

This constraint shows that **RTypeInference** thinks `y_1` is the return value
(since it is the value of the last expression). The type inference system
generates another type variable, `t2`, as a placeholder for the type of `y_1`
in the return call. The earlier assignment `y_1 <- x_1` established that `t2`
must have the same type as `t1`, so the inference system also generates this
constraint.

Notice that the constraint is not an `Equivalence` but instead an
`ImplicitInstance`. This kind of constraint allows for polymorphism. From the
perspective of `constrain`, the type variable `t1` is an unknown type, to be
established later on by solving for the type variables. As a consequence, `t1`
could be a polymorphic function type. In that case, `t2` is an instance of `t1`
(instantiated with concrete argument types, to be inferred based on how `y_1`
is called). This distinction between an `Equivalence` and an `ImplicitInstance`
is only important for function types; if `t2` is not a function type, then the
inference system will treat the `ImplicitInstance` the same way it would an
`Equivalence`. 

The result from `constrain` is an intermediate result in type inference. We've
presented it here because it is important to understand for anyone that wants
to tinker with the type inference system, and because even on its own it
contains useful information, such as where program variables are used.

To get to the final result from type inference, we need to have
**RTypeInference** solve the system of constraints. At a high level, the
strategy to solve the system is to treat each constraint as a substitution:
given a constraint, the type variables on one side can be replaced with the
types (or type variables) on the other side in all other constraints, and the
original constraint removed. Repeating this process eventually yields a
composition of substitutions that "solves" the constraint system, so that all
type variables equivalent to a concrete type are replaced by that type.

The **RTypeInference** function `csolve` computes the substitution that solves
the type inference problem. This function should be called directly on the
result from `constrain`. For instance, in the example:
```{r}
sub = csolve(ans)
sub
```

The result is a **typesys** `Substitution` object, which records the
equivalence between each type variable and type. Recall that the `SymbolMap` in
the result from `constrain` is our mapping between program variables and type
information. Thus to relate the substitution from `csolve` back to the program
variables in the source code, we need to call the substitution on the
`SymbolMap`. For convenience, we can also call the substitution on the entire
`Result` object. Here's the code:
```{r}
# sub(ans@map)
# OR
sub(ans)
```

The substituted `SymbolMap` is the final result from type inference. In this
case, inference is able to determine that `x_1`, `y_1`, and the return value
are all numeric.

This section described the basic workflow for **RTypeInference**: call
`constrain` to collect type information from the code, call `csolve` to use the
constraints to solve for the type variables, and finally, apply the resulting
substitution to the `SymbolMap` from `constrain` to relate the type variables
back to the program variables.


## A More Sophisticated Example


Now let's look at an example with a variable whose type is not made obvious
from an assignment. The code is a function that computes the area of an
annulus:
```{r}
circle_area = function(r_inner, r_outer) {
  outer = pi * r_outer^2
  inner = pi * r_inner^2

  return (outer - inner)
}
```

We can use the `to_blocks` function to convert the function code to SSA form
(rather than retyping the whole thing for `quote_blocks`). Then we can call
`constrain` as before:
```{r}
ex = to_blocks(circle_area)
ans = constrain(ex)
ans
```

As in the previous example, the result is a `Result` object with slots
`constraints` and `map`. All symbols in the source code, including names of
functions, are recorded in the symbol map, so it has many more entries here.
Variables in the global environment, like `*`, `pi`, and `^` here, are recorded
as having no definition in the provided source code.

Note: A bug in the current version of **RTypeInference** causes the parameters
to be left out of the `SymbolMap`. This is related to trying to support nested
function definitions and multiple scopes in type inference. I *think* the right
way to handle this is to have the constraint generator create a new `SymbolMap`
for each nested function, with a parent field (mimicking R's environments), but
haven't tested this.

In order to make type inference more accurate, we can provide the `constrain`
function with type information about variables up front (a future version of
the package could optionally inspect an environment to collect some of this
information automatically, or have a database of type information for built-in
functions).

For instance, suppose we want to specify that `pi` is a numeric value. We can
do this by creating a `SymbolMap` before calling `constrain`, and passing the
map to `constrain`. Here's the code to create the symbol map and add a
definition for `pi`:
```{r}
m = SymbolMap()
m = set_defined_as(m, "pi", RNumeric)
```

The `set_defined_as` function assigns a type to a specific program variable in
the `SymbolMap`. The corresponding `get_defined_as` retrieves the type a
program variable is defined as from a `SymbolMap`.

With the map defined, we can pass this information into the type inference
system when we call `constrain`:
```{r}
constrain(ex, m)
```

Similarly, we might want let the type inference system know that (in this case)
`*` is a function that takes two numeric values and returns a numeric value. We
can do that by adding another entry to the symbol map before calling
`constrain`:
```{r}
m = set_defined_as(m, "*", RFunction(RNumeric, RNumeric, RNumeric))
```

The **typesys** `RFunction` constructor creates a function type; the last
argument is always treated as the return type and the others as the argument.
In the current version of **typesys**, argument types are only matched
positionally, not by name. Let's generate the constraints with the updated
symbol map:
```{r}
ans = constrain(ex, m)
```

Although we haven't provided complete information to the type inference system
(for instance, we did not provide a type for `^`), the system can still recover
a lot of information from the code and what we did provide. Let's solve for the
type variables to see what we get:
```{r}
sub = csolve(ans)
sub(ans)
```

In the result, we can see that the system determined that `^` is a function,
and that its second argument and result must be numeric. The system was unable
to determine the type of the first argument to `^`, because the parameters
`r_outer` and `r_inner` could potentially have any type.
