---
title: "Type Inference Improvements"
author: "Nick Ulle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Type Inference Improvements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(RTypeInference)
```

## Multiple Functions

The current version of the package doesn't move between functions
automatically, so functions which call other functions in the same source file
are effectively unsupported. This is an essential feature since idiomatic R
code breaks tasks into short, composable functions.

### Search Path

The first issue is where the package should look for names.

For interactive use, it's convenient to search for names in the global
environment. 

On the other hand, scripts have their own global environment when evaluated. If
inference is run on a script from an interactive session, searching in the
global environment might be unexpected behavior. However, it would be useful
for on-the-fly modifications to scripts to test how inference responds.

Also note that inference may need to work across several scripts, so the
scripts will need to be loaded into a shared environment.

An alternative to the above is to load scripts with `parse()`. Scripts may
contain code that runs at eval, which probably shouldn't run during type
inference. Using `parse()` avoids evaluation during inference. However, this
also means the inference package is responsible for tracking down every name
that's loaded by the script, e.g., using `source()`.

### Types Data Structure

The second issue is what data structure the package should use for keeping
track of types for multiple scopes. Scopes have a tree-like hierarchy and this
must be maintained in order to handle closures correctly.

The simple solution is to use a named list of lists and TypeCollector objects.



## Type Handler API

Occasionally the user will want to define their own rules for how a function
should be typed. A custom _handler function_ can be provided for each function
where the default inference rules should be overridden. The handler function
must fulfill a contract with the type inference package in order to work
correctly.

The current version's function table only holds types, so handler functions are
wrapped in ConditionalType objects. 

## Rules

The rules used for inference are written in R. Some rules, such as those for
mathematical operators, are built in and applied as part of the AST-walk.
However, most rules are stored as separate hand-written R functions that return
a named list of types.

```{r, eval = FALSE}
  "abs" = ConditionalType(
    # complex|numeric -> numeric
    # integer -> integer
    function(args) {
      # Here x should be an vector, not a list.
      atom = element_type(args$x)

      atom =
        if (is(atom, "ComplexType") || is(atom, "RealType"))
          RealType()
        else if (is(atom, "IntegerType"))
          IntegerType()

      element_type(args$x) = atom
      # FIXME:
      value(args$x) = UnknownValue()
      return(args$x)
    }),
```

This makes it difficult to make inferences based on combinations of rules,
because there is no procedure for combining rules that cannot already be
resolved to a type. On the other hand, this allows allows arbitrarily complex
rules to be encoded (resolving some rules could take as long as running the
input program).

There are several ways to address this. First, we could define a grammar for
rules and build a simple parser. This leads to a formal system for resolving
rules and might allow us to prove that our system is sound. However, it's not
as flexible as the current system. Another advantage of this approach is that
it might make automatic generation of rules comparatively easy.

Second, we could use a system of attributes/tags on functions of interest to
generate rules implicitly. 

Third, we could preserve the current system.

## 

## Lists & Data Frames

At present RTI can't infer any information about data frames.

```{r}
df_fun = function() {
  data.frame()
}

tc = TypeCollector()
inferTypes(df_fun, tc)
```

The real issue here is that RTI doesn't even try to detect data frames. So the
first step towards a fix is to actually recognize data frames. This could be
based on `data.frame` and also how `$` is used.

